---
title: "RevisedNon470"
output: bookdown::html_document2
date: "2023-03-03"
---

```{r Front Matter, include=FALSE}
##set up
rm(list = ls())
library(tinytex)
library(tidyverse)
library(data.table)
library(caret) #For creating confusion matrix and associated probabilities
library(pROC)
library(car)
library(mosaic)
library(lme4)

a<-"~/Desktop/PSU/SeniorYear/FallSemester/STAT400/STAT400_FinalProj/AlgerianForests.csv"
forests <- fread(a)
forests <- as.data.frame(forests)
#add dates
forests$date <- as.Date(with(forests,paste(year,month,day,sep="-")),"%Y-%m-%d")
#code for region indicators
forests$regionBejai <- ifelse(forests$Region == "Bejaia",1,0)
#code indicator for fire
forests$FireNumb <- ifelse(forests$Classes == "fire",1,0)
#omit unnecessary variables
forest <- as.data.table(forests)
a <- c("Temperature","RH","Ws","Rain","regionBejai","FireNumb")
forestfinal <- forest[,..a]
#rename cols
colnames(forestfinal) <- c('Temp','RelHumid','WindSpd','TotalRain','RegionBejaia',"FireNumb")
#rename data to be used for analysis
fdata <- forestfinal
fdata$FireNumb <- as.factor(fdata$FireNumb)
fdata$RegionBejaia <- as.factor(fdata$RegionBejaia)
```

# Forest Fires in Algeria


# Introduction 

The following analysis serves as a general overlook as to the usage of logistic regression to predict instances of forest fires occurring in Algeria. This is mean to serve as a general analysis but the information contained is useful for individuals in the field of fire protection as to ensure that there is a broader awareness as to how specific weather parameters play into the odds of a new fire emerging. 

In the case in which we are looking at the forest fires in Algeria, researchers account for instances in which a new forest fire has occurred in a given region based on the fire weather index system (FWI). In turn this means that in an instance in which a fire in a specified area of the region continuously fulfills the FWI criteria for a fire and burns for several days, it is not accounted for in the data. Rather the data aims to classify new blazes occurring in the region on a given day, researchers used this as a facet to eliminate possible correlation in the data (Abid & Izeboudjen 2019). The original data is composed of 18 variables from those 18 variables, we refined to only look at 6 total variables, 1 response with 5 features utilized to predict the odds of the response being equal to 1.

From our refined data set we aim to utilize logistic regression to classify an instance as a forest fire or not based on field readings of temperature, relative humidity, wind speed, total rain and region. The variables are coded as follows:


```{r,echo=FALSE}
Variables <- colnames(fdata)
Ranges <-c("22-42","21-90","6-29","0-16.8","0 or 1","0 or 1")
Description<-c("Temperature in degrees Celsius","Relative Humidity (percentage 0-100)","Total rain fall in millimeters","1 for region is Bejaia, 0 for Sidi-Bel Abbes","Wind Speed in mph","1 for a forest fire, 0 for no forest fire occurring")
VarTab<-cbind(Variables,Description,Ranges)
knitr::kable(VarTab,caption = "")
```

It is well understood that in hotter drier days more fires are expected to occur (CIRES 2022). In turn our base model will include those two features and more features will be added in effort to improve the model.


Research Questions: 

What features should a model include to best explain the variability in instances of new forest fires emerging on a given day in the Algerian Forests in the Bejaia & Sidi-Bel Abbes Regions?

How accurately can we predict the risk of a future fire emerging on a given day in the Algerian Forests in the Bejaia & Sidi-Bel Abbes Regions?


# Exploratory Data Analysis

## Prepare Data for Analysis
```{r}
a<-"~/Desktop/PSU/SeniorYear/FallSemester/STAT400/STAT400_FinalProj/AlgerianForests.csv"
forests <- fread(a)
forests <- as.data.frame(forests)
#add dates
forests$date <- as.Date(with(forests,paste(year,month,day,sep="-")),"%Y-%m-%d")
#code for region indicators
forests$regionBejai <- ifelse(forests$Region == "Bejaia",1,0)
#code indicator for fire
forests$FireNumb <- ifelse(forests$Classes == "fire",1,0)
#omit unnecessary variables
forest <- as.data.table(forests)
a <- c("Temperature","RH","Ws","Rain","regionBejai","FireNumb")
forestfinal <- forest[,..a]
#rename cols
colnames(forestfinal) <- c('Temp','RelHumid','WindSpd','TotalRain','RegionBejaia',"FireNumb")
#rename data to be used for analysis
fdata <- forestfinal
fdata$FireNumb <- as.factor(fdata$FireNumb)
fdata$RegionBejaia <- as.factor(fdata$RegionBejaia)
```


The data set was modified creating 2 dummy variables for forest fires where 0 is no fire and 1 represents a fire, and for region being Bejaia where 0 represents a region other than Bejaia and 1 represents Bejaia. Additionally the field reading variables are selected due to their easier interpretability and weight in determining other meteorological variables in the data


## Frequency Tables

```{r}
# frequency of fires vs no fire per region
Table1 <- plyr::count(fdata,"FireNumb")
Nofire <- fdata %>% 
  select(FireNumb,RegionBejaia) %>% 
  filter(FireNumb==0)
Fire <- fdata %>% 
  select(FireNumb,RegionBejaia) %>% 
  filter(FireNumb==1)
Nofirefreq <- plyr::count(Nofire,"RegionBejaia")
Firefreq <- plyr::count(Fire,"RegionBejaia")
FreqTable <- inner_join(Nofirefreq,Firefreq, by = "RegionBejaia")%>%rename(FreqNoFire = "freq.x",FreqFire = "freq.y")
```
```{r, echo=FALSE}
knitr::kable(Table1,caption = "Frequencies of Fires and No Fires in the Data Set")
```
```{r, echo=FALSE}
knitr::kable(FreqTable,caption = "Frequencies of Fires and No Fires per Region")
```


The frequency tables demonstrates that in our data set there are 138 instances of a given fire being classified as a new fire. The data also shows from the new fires that occurred that most happened in the Sidi-Bel Abbes Region, and proportionally speaking more new fires occur in the Sidi-Bel Abbes Region.


## Plots
```{r ,fig.cap = "Frequency of fire versus no fire for each region", echo= FALSE}
 ggplot(fdata) +
  aes(x = FireNumb, fill = FireNumb) +
  geom_bar() +
  scale_fill_hue(direction = 1) +
  theme_minimal() +
  facet_wrap(vars(RegionBejaia))
```


The bar graph demonstrates that most of the fires occur in the Sidi-Bel Abbes, and that from a proportion standpoint more instances of no new emerging fires occur in the Bejaia region. 


```{r, fig.cap="Scatter for possible multi collinearity", echo= FALSE}
a1 <- ggplot(fdata) +
  aes(x = RelHumid, y = TotalRain, colour = FireNumb) +
  geom_point(shape = "circle", size = 1.5) +
  scale_color_hue(direction = 1) +
  theme_minimal()
a2 <- ggplot(fdata) +
  aes(x = RelHumid, y = Temp, colour = FireNumb) +
  geom_point(shape = "circle", size = 1.5) +
  scale_color_hue(direction = 1) +
  theme_minimal()
gridExtra::grid.arrange(a1,a2, ncol = 2)
```


The scatter plots that there might be some slight correlation between the variables relative humidity and total rain, we see a positive trend occur in the scatter plot showcasing a positive correlation, the plot also demonstrates that with higher humidity and total rain less new fires occur. The scatter plot for temperature and relative humidity demonstrates a negative correlation, the plot also demonstrates that with a larger temperature and less humidity more new fires occur. These instances of correlation between variable could lead to multicollinearity in a model which utilizes both features resulting in a model that does not adequately explain the observed variability.


```{r, fig.cap="Boxplots Looking at Distributions", echo= FALSE}

p1 <- ggplot(fdata) +
  aes(x = RelHumid, y = FireNumb, fill = FireNumb) +
  geom_boxplot() +
  scale_fill_hue(direction = 1) +
  theme_minimal()
p2 <- ggplot(fdata) +
  aes(x = TotalRain, y = FireNumb, fill = FireNumb) +
  geom_boxplot() +
  scale_fill_hue(direction = 1) +
  theme_minimal()
p3 <- ggplot(fdata) +
  aes(x = WindSpd, y = FireNumb, fill = FireNumb) +
  geom_boxplot() +
  scale_fill_hue(direction = 1) +
  theme_minimal()
p4 <- ggplot(fdata) +
  aes(x = Temp, y = FireNumb, fill = FireNumb) +
  geom_boxplot() +
  scale_fill_hue(direction = 1) +
  theme_minimal()
gridExtra::grid.arrange(p1,p2,p3,p4, ncol = 2)
```

In looking at the distribution of values for each of the variables in our data set it is noticeable that new fires occur at  higher temperatures, lower rainfall and lower relative humidity. The distribution for wind speed is relatively the same for instances of fire and no fires, this could indicate that wind speed doesn't play a role in new emerging fires btween the two regions.  



# Models 

In this case the reasoning for utilizing logistic regression is due to the binary classification of either a new fire being present or absent from the given region. Logistic regression utilizes the logit which allows for the log odds to have a linear relationship between parameters observed in the data set. For the predictive aspect in determining a fire being present or not a threshold of .5 is utilized any instance in probability of a fire above .5 is considered a new emerging fire. Logistic regression is an optimal method due to the independence established in the data by the researchers.

The main conditions necessary for the utilization of logistic regression are the independence condition and the response variable being binary. In this case the conditions for logistic regression are fulfilled since the response variable is the is the FireNumb which has 2 response either 0 or 1. In addition to that the condition for independence is fulfilled as each case of a fire is a new fire that is independent of a previous existing fire in the region.

```{r}
#create full model of data
full_mod <- glm(FireNumb ~ (.)^2 ,family = "binomial",data = fdata)
#warning indicates that one of these variables explains the variability observed in log odds to a very high degree
params <- summary(full_mod)$coefficients
```

```{r, echo=FALSE}
knitr::kable(params,caption = "Summary of Parameters for Full Model")
```


In building the initial model it is necessary to account for the most readily definable sources of variability in turn a model with all the features included and all 2 way interaction terms is utilized. By choosing this model as a starting point there is a smaller number of assumptions about what may cause an instance of a new emerging forest fire in the regions of Algeria.




## Feature Selection
```{r, results='hide',warning=FALSE}
#feature selection using both forward and backward elimination starting from the intercept only model.
#intercept only model
null <- glm(FireNumb ~ 1,data = fdata,family="binomial")
#both directions of stepwise regression
stats::step(null, 
            scope = list(upper=full_mod), 
            data = fdata, direction = 'both')
```


Utilizing forward selection demonstrates that the best model is the model which takes all the features in the data into account, however in looking at the change in AIC it is noticeable that after adding the total rain variable the drop in AIC is not very large. Although the change in the AIC is not very large the model with all the features should be utilized because there is still a drop in AIC which demonstrates that adding more features doesn't pose a penalty to the model's ability to explain the variability.

The best model follows the formula,

FireNumb ~ TotalRain + Temp + RelHumid + WindSpd + TotalRain:Temp + TotalRain:RelHumid + TotalRain:WindSpd + Temp:WindSpd



## Model Comparison
```{r}
#create reduced model found in both stepwise selection
reduced_mod <-glm(formula = FireNumb ~ TotalRain + Temp + RelHumid + WindSpd + 
    TotalRain:Temp + TotalRain:RelHumid + TotalRain:WindSpd + Temp:WindSpd, 
    family = "binomial", data = fdata)
#drop in deviance test for 
ANO <- anova(reduced_mod,full_mod, test = "Chisq")
```

```{r}
#goodness of fit test pvalue
x2<-sum(residuals(reduced_mod,type = "pearson")^2)
pval<-1-pchisq(reduced_mod$deviance,reduced_mod$df.residual)
pval
```

```{r, echo = FALSE}
knitr::kable(ANO,caption = "Drop-in-Deviance Test for Reduced vs Full Model")
```



$$H_o: \beta_j = 0$$

$$H_a: \beta_j \neq 0$$ 


In comparing the initial model versus the model optimized through step selection we observe that there is evidence to suggest not incluiding all the features improves the model, meaning the reduced model for emerging fires is better at explainignt he observed variability than the full model with all the two way interactions. Similarly this also suggests that the slope of at least one of the newly added features is 0. This is further proved when looking at the test statistic which is 3.840707 with and associated p-value of 0.7979313 which is greater than .05. The goodness of fit tests also has an associated value of 1 which implies that there is insufficient evidence to suggest the reduced model does not fit the data set.



## Best Model Interpretation
```{r, echo=FALSE,warning=FALSE}
b <- summary(reduced_mod)
coeff <- b$coefficients
CIS <-confint(reduced_mod, level=0.95)
coeff <- cbind(coeff,CIS)
knitr::kable(coeff,caption = "Parameter Summaries and Confidence Intervals for the Best Model")
```


The confidence interval demonstrates that the significant features which do not hold zero within their confidence interval are total rain variable, relative humidity, the interaction between totalrain and relhumid, interaction between total rain and windspeed, and interaction of totalrain and temp. The interpretation for the most significant non-interaction feature relative humidity variable is that for every one unit increase in relative humidity there is 95% confidence that the odds of a new emerging fire decreases by 0.8856661 to 0.9840544.



The best model has 8 total parameters Temp, RelHumid, WindSpd, TotalRain, RegionBejaia, TotalRain:Temp, TotalRain:WindSpd, TotalRain:WindSpd and Temp:WindSpd. The feature for temperature is the least significant however its interaction term is significant, therefore it is kept in the model to improve its interpretability. To explain the coefficients in the model when all other features are held constant and there is a single unit increase in a given feature the corresponding estimate means there is an increase or decrease by the estimate amount to the log odds, if this value is exponentiated then it show casesby how much of a factor the odds is altered. To exemplify this concept when looking at total rain for every one unit increase in total rain it is expected the log odds of a new fire emerging decreases by 143.0829849 and the odds of a new fire emerging to decrease by a factor of $e^{-143.0829849}$(7.241845e-63).


# Classification Analysis
```{r}
#set a threshold
threshold <- 0.5
#predict using full model
pred_prob <- predict(reduced_mod, newdata = fdata, type = "response")
fdata$pred_prob <- pred_prob
pred_10 <- ifelse(pred_prob > threshold, 1, 0)
#add feature distinguishing fire based on threshold
fdata$pred_10 <- ifelse(pred_prob > threshold, 1, 0)
#confusion matrix for forest fire
CM <- confusionMatrix(data = as.factor(pred_10), reference = fdata$FireNumb, positive = "1")
```

```{r,echo=FALSE}
knitr::kable(CM$table,caption = "Confusion Matrix for the best Model")
```

```{r,echo=FALSE}
Evals <- as.data.frame(CM$byClass)
Evals$values <- Evals$`CM$byClass`
Evals$`CM$byClass`<-NULL
x<-data.frame("Accuracy",values = 0.918)
rownames(x)<-"Accuracy"
x$X.Accuracy.<-NULL
Evals<-rbind(x,Evals)
knitr::kable(Evals,caption = "Evaluation Metrics")
```

The model has an overall accuracy of .918, which showcases that the model is effective at predicting instances of fires given the features in the data. The sensitivity of the model is .9637681 and the specificity is .8584906, this shows that the model is more effective at distinguishing an instance of a new emerging fire rather than an instance where a new fire has not emerged. 


```{r,fig.cap="ROC curve of full model with AUC = .965"}
test_roc = roc(response = fdata$FireNumb, predictor = pred_prob, plot = TRUE, print.auc = TRUE, legacy.axes=TRUE)
```


The AUC for the model is .965 which is close to one which shows that the model is effective at classifying instances of new emerging fires, meaning the model is performs better than random chance.


# Conclusion
In conclusion the analysis has demonstrated that including all the features and their interactions present in the refined data set does not create a better model than the model selected by stepwise selection in both directions which only accounts for relative humidity and temperature. This is demonstrated by the usage of both directions of feature selection results in the best model was found to be the model which includes only Temp, RelHumid, WindSpd, TotalRain, RegionBejaia, TotalRain:Temp, TotalRain:WindSpd, TotalRain:WindSpd and Temp:WindSpd. In turn this is also further justified when looking at the drop-in-deviance test which demonstrated a test statistic of 3.840707 with a associated p-value of 0.7979313, the p-value was larger than .05 which leads to support the conclusion that the model is improved upon only selecting a subset of features and interactions in the data. The model's adequacy is further justified when looking at the goodness of fit test which demonstrates that the new model does fit the data, since its associated p-value is at 1 which is greater than .05, meaning the null is not rejected meaning there is insufficient evidence to suggest the model doesn't fit the data. 

The analysis also demonstrates that the best model is able to predict the instances of new emerging fires to a better capacity than random chance is capable of doing. The accuracy of the model is roughly .918 which is larger than .5 meaning that the model is able to predict an instance of a new emerging fire to a capacity in which the model is better than utilizing random chance. The ROC curve for the model also demonstrates that the model is good at classification, since the AUC value is .965 which is close to 1, this also supports the claim that the model performs better than random chance.

One major drawback of this analysis can be seen in the predictive power of the model. When generating the the predicitons it is observed that there are a few instances where the probability becomes equal to one. This could imply that our model may have some overfitting, where it rather explains the observed variability in the data rather than the relationship at large between variables. In order to solve this it could be useful to facilitate an analysis that also utilizes cross validation as a component of it's process. To combat the plausibility of over fitting, an approach utilizing cross validation could prove more useful in finding th optimal model to best explain new emerging instances of forest fires in the Algerian forests.

Works Cited

Hamadeh, N., Karouni, A., & Daya, B. (2014). Predicting Forest Fire Hazards Using Data Mining Techniques: Decision Tree and Neural Networks. Advanced Materials Research, 1051, 466–470. https://doi.org/10.4028/www.scientific.net/amr.1051.466

N/A (2022, February 10). Hotter, Drier Nights Mean More Runaway Fires. CIRES. https://cires.colorado.edu/news/hotter-drier-nights-mean-more-runaway-fires

# Appendix 

Additional Summary Statistics

```{r}
#stats for numerical variables comparing instances of fire versus non-fire
f1 <- favstats(RelHumid ~ FireNumb, data = fdata)
f1$id<-"RelHumid"
f2 <- favstats(TotalRain ~ FireNumb, data = fdata)
f2$id<-"TotalRain"
f3 <- favstats(WindSpd ~ FireNumb, data = fdata)
f3$id<-"WindSpd"
f4 <- favstats(Temp ~ FireNumb, data = fdata)
f4$id<-"Temp"
tab <- rbind(f1,f2,f3,f4)

#stats for each numerical variable
a <- favstats(fdata$RelHumid) %>% mutate(id = "RelHumid")
b <- favstats(fdata$TotalRain)  %>% mutate(id = "TotalRain")
c <- favstats(fdata$WindSpd)  %>% mutate(id = "WindSpd")
d <- favstats(fdata$Temp) %>% mutate(id = "Temp")
e<-rbind(a,b,c,d)
table <- e[,c("id","mean","sd","min","median")]
```


```{r, echo=FALSE}
knitr::kable(table)
```

The mean for relative humidity is 61.93 with a standard deviation of 14.884200, this means that the two regions remain relatively largely humid. The total rain mean is 0.7606557 which demonstrates that the regions do not get very much rain throughout the year. The mean wind speed is 15.5 which shows that the region is not particularly windy. The mean temperature is 32.1721311 which shows that the regions are relatively warm throughout the year.


```{r, echo=FALSE}
knitr::kable(tab,caption = "Summary Stats of Weather Parameters in Different Fire Instances")
```

The stats between instances when a  new fire occurs versus when a new fire doesn't occur demonstrates that the mean value of temperature is larger for when a new fire occurs, the stats also demonstrate that the total rain and relative humidity have smaller means when a fire occurs, and the stats demonstrate that the difference in wind speed is relatively small between when a fire occurs or does not occur. This demonstrates that there may be some sort of association for a new fire to occur at higher temperatures, smaller total humidity and smaller rain fall.

